
---
title: "IODS final project"
author:
  name: Riku Laine
  affiliation: riku.laine@helsinki.fi
output:
  html_document:
    theme: yeti
    toc: true
    toc_depth: 3
    fig_caption: true
    code_folding: hide
---

<center> <h1>PROJECT TITLE</h1> </center>

```{r Settings, include=FALSE}
library(knitr)
library(ggplot2)
library(GGally)
library(corrplot)
opts_chunk$set(message = F)
opts_chunk$set(warning = F)
opts_chunk$set(out.width = "100%")

mycor <- function(data, mapping, alignPercent = 0.6, method = "pearson", 
    use = "complete.obs", corAlignPercent = NULL, corMethod = NULL, 
    corUse = NULL, sgnf=3, ...) {
    if (!is.null(corAlignPercent)) {
        stop("'corAlignPercent' is deprecated.  Please use argument 'alignPercent'")
    }
    if (!is.null(corMethod)) {
        stop("'corMethod' is deprecated.  Please use argument 'method'")
    }
    if (!is.null(corUse)) {
        stop("'corUse' is deprecated.  Please use argument 'use'")
    }
    useOptions <- c("all.obs", "complete.obs", "pairwise.complete.obs", 
        "everything", "na.or.complete")
    use <- pmatch(use, useOptions)
    if (is.na(use)) {
        warning("correlation 'use' not found.  Using default value of 'all.obs'")
        use <- useOptions[1]
    } else {
        use <- useOptions[use]
    }
    cor_fn <- function(x, y) {
        cor(x, y, method = method, use = use)
    }
    xCol <- deparse(mapping$x)
    yCol <- deparse(mapping$y)
    if (GGally:::is_date(data[[xCol]]) || GGally:::is_date(data[[yCol]])) {
        if (!identical(class(data), "data.frame")) {
            data <- fix_data(data)
        }
        for (col in c(xCol, yCol)) {
            if (GGally:::is_date(data[[col]])) {
                data[[col]] <- as.numeric(data[[col]])
            }
        }
    }
    if (is.numeric(GGally:::eval_data_col(data, mapping$colour))) {
        stop("ggally_cor: mapping color column must be categorical, not numeric")
    }
    colorCol <- deparse(mapping$colour)
    singleColorCol <- ifelse(is.null(colorCol), NULL, paste(colorCol, 
        collapse = ""))
    if (use %in% c("complete.obs", "pairwise.complete.obs", "na.or.complete")) {
        if (length(colorCol) > 0) {
            if (singleColorCol %in% colnames(data)) {
                rows <- complete.cases(data[c(xCol, yCol, colorCol)])
            } else {
                rows <- complete.cases(data[c(xCol, yCol)])
            }
        } else {
            rows <- complete.cases(data[c(xCol, yCol)])
        }
        if (any(!rows)) {
            total <- sum(!rows)
            if (total > 1) {
                warning("Removed ", total, " rows containing missing values")
            } else if (total == 1) {
                warning("Removing 1 row that contained a missing value")
            }
        }
        data <- data[rows, ]
    }
    xVal <- data[[xCol]]
    yVal <- data[[yCol]]
    if (length(names(mapping)) > 0) {
        for (i in length(names(mapping)):1) {
            tmp_map_val <- deparse(mapping[names(mapping)[i]][[1]])
            if (tmp_map_val[length(tmp_map_val)] %in% colnames(data)) 
                mapping[[names(mapping)[i]]] <- NULL
            if (length(names(mapping)) < 1) {
                mapping <- NULL
                break
            }
        }
    }
    if (length(colorCol) < 1) {
        colorCol <- "ggally_NO_EXIST"
    }
    if ((singleColorCol != "ggally_NO_EXIST") && (singleColorCol %in% 
        colnames(data))) {
        cord <- plyr::ddply(data, c(colorCol), function(x) {
            cor_fn(x[[xCol]], x[[yCol]])
        })
        colnames(cord)[2] <- "ggally_cor"
        cord$ggally_cor <- signif(as.numeric(cord$ggally_cor), 
            sgnf)
        lev <- levels(data[[colorCol]])
        ord <- rep(-1, nrow(cord))
        for (i in 1:nrow(cord)) {
            for (j in seq_along(lev)) {
                if (identical(as.character(cord[i, colorCol]), 
                  as.character(lev[j]))) {
                  ord[i] <- j
                }
            }
        }
        cord <- cord[order(ord[ord >= 0]), ]
        cord$label <- GGally:::str_c(cord[[colorCol]], ": ", cord$ggally_cor)
        xmin <- min(xVal, na.rm = TRUE)
        xmax <- max(xVal, na.rm = TRUE)
        xrange <- c(xmin - 0.01 * (xmax - xmin), xmax + 0.01 * 
            (xmax - xmin))
        ymin <- min(yVal, na.rm = TRUE)
        ymax <- max(yVal, na.rm = TRUE)
        yrange <- c(ymin - 0.01 * (ymax - ymin), ymax + 0.01 * 
            (ymax - ymin))
        p <- ggally_text(label = GGally:::str_c("Corr: ", signif(cor_fn(xVal, 
            yVal), sgnf)), mapping = mapping, xP = 0.5, yP = 0.9, 
            xrange = xrange, yrange = yrange, color = "black", 
            ...) + theme(legend.position = "none")
        xPos <- rep(alignPercent, nrow(cord)) * diff(xrange) + 
            min(xrange, na.rm = TRUE)
        yPos <- seq(from = 0.9, to = 0.2, length.out = nrow(cord) + 
            1)
        yPos <- yPos * diff(yrange) + min(yrange, na.rm = TRUE)
        yPos <- yPos[-1]
        cordf <- data.frame(xPos = xPos, yPos = yPos, labelp = cord$label)
        cordf$labelp <- factor(cordf$labelp, levels = cordf$labelp)
        p <- p + geom_text(data = cordf, aes(x = xPos, y = yPos, 
            label = labelp, color = labelp), hjust = 1, ...)
        p
    }  else {
        xmin <- min(xVal, na.rm = TRUE)
        xmax <- max(xVal, na.rm = TRUE)
        xrange <- c(xmin - 0.01 * (xmax - xmin), xmax + 0.01 * 
            (xmax - xmin))
        ymin <- min(yVal, na.rm = TRUE)
        ymax <- max(yVal, na.rm = TRUE)
        yrange <- c(ymin - 0.01 * (ymax - ymin), ymax + 0.01 * 
            (ymax - ymin))
        p <- ggally_text(label = paste("Corr:\n", signif(cor_fn(xVal, 
            yVal), sgnf), sep = "", collapse = ""), mapping, xP = 0.5, 
            yP = 0.5, xrange = xrange, yrange = yrange, ...) + 
            theme(legend.position = "none")
        p
    }
}
```

## Abstract

Some stuff $$y = \alpha x_1 + \beta x_2 + \epsilon$$, which happens to be regression function for linear regression.

## Research question and hypothesis

I was interested in finding out what factors of people's learning are the most connected to their final score. Could boundness in the course syllabus be somehow in connection with a low score? I concluded that logistic regression was the most fitting tool to answer this question.

I hypothesized that -- along with attitude -- `RelatingIdeas` would affect the final score as it represents how much does a atudent ponder the things taught in their free time. And when one revises things learned it deepens their learning, as "repetitio est mater studiorum".

## Data description

[The data](http://www.helsinki.fi/~kvehkala/JYTmooc/JYTOPKYS3-data.txt) was acquired from a study made by Kimmo Vehkalahti. The study included a questionnaire which was held in Fall 2014 for the participants of an introductory course to statistics (Johdatus yhteiskuntatilastotieteeseen) and it was funded by the Teachers' Academy. The questionnaire itself had questions regarding for instance the students' motivation, strategies in studying and general interest in statistics.

### Data processing

The original data set consisted of 183 observations, but after [processing it](https://raw.githubusercontent.com/Riku-Laine/IODS-final/master/data/createLearningFP.R) only 166 remained from the exclusion of individuals who had zero points from the final exam from not attending it. Processing the data, I constructed the variables for e.g. use of evidence and seeking meaning by the instructions in the [meta page](http://www.helsinki.fi/~kvehkala/JYTmooc/JYTOPKYS2-meta.txt) and named them to be self explanatory. For instance, `SeekingMeaning` is the mean for the answers D03, D11, D19 and D27 which were answered in the 5-point [Likert scale](https://en.wikipedia.org/wiki/Likert_scale).

### The data visually

*Visually clear and interesting explorations of the variables of interest in the data, from the point of view of your research question. Include interpretations of the distributions and relationships of the variables. Use captions to draw the reader’s focus on the interesting parts of your tables and graphics.*

First I explored the data with a paired plot matrix to see possible trends and correlations between the variables of the data. Afterwards, I deepened my analysis with some variables of interest revealed by the plot matrix.

```{r Plots}
learning2014 <- read.csv("~/GitHub/IODS-final/data/learning2014.csv")

ggpairs(learning2014, mapping = aes(col=Gender, alpha=0.1), lower = list(combo = wrap("facethist", bins = 20)),  upper = list(continuous = wrap(size = 1.7, mycor, sgnf=2)))+
	theme(strip.text = element_text(size=6), 
				strip.text.y = element_text(angle=0), 
				axis.text = element_text(size=6), 
				strip.text.x = element_text(angle=45))+
	ggtitle("Figure 1: Interactions in Learning2014 data")
```

Distributions of the variables seem to be quite symmetrical, although most of the respondents are students and therefore it is not surprising that the inter-quartile range of age falls between 21 and 27 years. There are only minor differences between the genders, males seem to have a better attitude towards statistics as females appear to be more organized in their studies. Males' better attitude towards mathematics have been shown in previous studies.

From figure 1 above it is visible that there are some quite strong correlations with some of the variables. For instance the variables of deep learning `OrganizedStudying` and `TimeManagement` have a correlation coefficient of $\rho \approx 0,7$ as the attitude of a student correlates with their points with a coefficient $\rho \approx 0,43$. Correlations are analyzed more thoroughly below figure 2.

```{r Corrplot, out.height="80%", out.width="80%"}
corrplot(cor(learning2014[, -1]), type = "lower", diag = T, tl.col='black', tl.srt = 0.1, tl.offset = 1, 
				 title = "Fig. 2: Correlations of learning2014 data")
```

From figure 2 it can be asserted that our variables correlate the most with the variables they comprise the original variables (such as Deep_adj) with. Also notable is the correlation with the students' attitude and points from the final exam being approximately `r round(cor(learning2014$Points, learning2014$Attitude), digits=2)`. This means better attitude associates with better performance in the exam. There are practically no other correlations that could cause collinearity in our regression model as all the others have an absolute value approximately smaller than 0.3.

### Gender differences in level of organization in studying and attitude toward statistics

In figure 1 there were some gender differences noted. Let's look briefly into them. First, let us construct boxplots of those variables and do a quick statistical test to assert the observed differences.

```{r plotteja, fig.asp=.4}
multiplot <- function(..., plotlist=NULL, file, cols=1, layout=NULL) {
  library(grid)

  # Make a list from the ... arguments and plotlist
  plots <- c(list(...), plotlist)

  numPlots = length(plots)

  # If layout is NULL, then use 'cols' to determine layout
  if (is.null(layout)) {
    # Make the panel
    # ncol: Number of columns of plots
    # nrow: Number of rows needed, calculated from # of cols
    layout <- matrix(seq(1, cols * ceiling(numPlots/cols)),
                    ncol = cols, nrow = ceiling(numPlots/cols))
  }

 if (numPlots==1) {
    print(plots[[1]])

  } else {
    # Set up the page
    grid.newpage()
    pushViewport(viewport(layout = grid.layout(nrow(layout), ncol(layout))))

    # Make each plot, in the correct location
    for (i in 1:numPlots) {
      # Get the i,j matrix positions of the regions that contain this subplot
      matchidx <- as.data.frame(which(layout == i, arr.ind = TRUE))

      print(plots[[i]], vp = viewport(layout.pos.row = matchidx$row,
                                      layout.pos.col = matchidx$col))
    }
  }
}
p1 <- ggplot(learning2014, aes(x=Gender, y=Attitude, fill=Gender))+
	geom_boxplot()+
	geom_jitter(position=position_jitter(0.1), alpha=.4)+ggtitle("Gender differences", subtitle = "Fig. 3 Attitude")

p2 <- ggplot(learning2014, aes(x=Gender, y=OrganizedStudying, fill=Gender))+
	geom_boxplot()+
	geom_jitter(position=position_jitter(0.1), alpha=.4)+ggtitle("", subtitle = "Fig. 4 OrganizedStudying")

multiplot(p1, p2, cols=2)
```

It can be seen that men have higher attitude points, which is supported by Kruskal-Wallis test ($p \approx 1.1 \cdot 10^{-4}$). Females seem to be more organized ($p \approx 7.1 \cdot 10^{-3}$) in a statistically significant quantity. These differences have been noted in previous studies **JATKA**

## Methods

### Logistic regression

As [NCBI](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC1065119/) puts it, logistic regression, "is a method for modelling the dependence of a binary response variable on one or more explanatory variables." In our case we have the binary response variable of points gotten in the final exam; if one has a score higher than **PÄÄTÄÄTTÄTÄ** they will have 1, as ones with a lower score get a zero. The regression model attemps to estimate the dependant variablein the best fashion. blaaa

### Other statistical tests used

#### Kruskal-Wallis test

Kruska-Wallis is a non-parametric alternative for the t-test. In other words, we do not have the assume that our variables are either homoscedastic -- to have the same variance -- or normally distributed.


## Results

### Summaries and visualizations

### Interpretations

### Validation analysis



## Conclusions and discussion
