
---
title: "IODS final project"
author:
  name: Riku Laine
  affiliation: riku.laine@helsinki.fi
output:
  html_document:
    theme: yeti
    toc: true
    toc_depth: 3
    fig_caption: true
    code_folding: hide
---

<center> <h1>PROJECT TITLE</h1> </center>

```{r Settings, include=FALSE}
library(knitr)
library(ggplot2)
library(GGally)
library(corrplot)
opts_chunk$set(message = F)
opts_chunk$set(warning = F)
opts_chunk$set(out.width = "100%")
```

## Abstract

Some stuff $$y = \alpha x_1 + \beta x_2 + \epsilon$$, which happens to be regression function for linear regression.

## Research question and hypothesis

I was interested in finding out what factors of people's learning are the most connected. Could boundness in the course syllabus be somehow in connection with student's time managing? I concluded that the principal component analysis (PCA) was the most fitting tool to answer this question. In addition, I performed logistic regression on some of the variables to find out which of the components affect the final score the most.

I hypothesized that the most similar would be the variables which make the final three analysis variables Deep_adj, Surf_adj and Stra_adj. For the logistic regression I presumed that seeking meaning, unrelated memorizing and attitude would affect the final score the most. 

## Data description

[The data](http://www.helsinki.fi/~kvehkala/JYTmooc/JYTOPKYS3-data.txt) was acquired from a study made by Kimmo Vehkalahti. The study included a questionnaire which was held in Fall 2014 for the participants of an introductory course to statistics (Johdatus yhteiskuntatilastotieteeseen) and it was funded by the Teachers' Academy. The questionnaire itself had questions regarding for instance the students' motivation, strategies in studying and general interest in statistics.

### Data processing

The original data set consisted of 183 observations, but after [processing it](https://raw.githubusercontent.com/Riku-Laine/IODS-final/master/data/createLearningFP.R) only 166 remained from the exclusion of individuals who had zero points from the final exam from not attending it. Processing the data, I constructed the variables for e.g. use of evidence and seeking meaning by the instructions in the [meta page](http://www.helsinki.fi/~kvehkala/JYTmooc/JYTOPKYS2-meta.txt) and named them to be self explanatory. For instance, `SeekingMeaning` is the mean for the answers D03, D11, D19 and D27 which were answered in the 5-point [Likert scale](https://en.wikipedia.org/wiki/Likert_scale).

### The data visually

*Visually clear and interesting explorations of the variables of interest in the data, from the point of view of your research question. Include interpretations of the distributions and relationships of the variables. Use captions to draw the reader’s focus on the interesting parts of your tables and graphics.*

First I explored the data with a paired plot matrix to see possible trends and correlations between the variables of the data. Afterwards, I deepened my analysis with some variables of interest revealed by the plot matrix.

```{r Plots, cache=TRUE}
learning2014 <- read.csv("~/GitHub/IODS-final/data/learning2014.csv")

ggpairs(learning2014, mapping = aes(col=Gender, alpha=0.3), lower = list(combo = wrap("facethist", bins = 20)))+
	theme(strip.text = element_text(size=6), axis.text = element_text(size=6))+
	ggtitle("Figure 1: Interactions in Learning2014 data")
```

Distributions of the variables seem to be quite symmetrical, although most of the respondents are students and therefore it is not surprising that the inter-quartile range of age falls between 21 and 27 years. There are only minor differences in between the genders, males seem to have a  better attitude towards statistics as females appear to be more organized in their studies. Mens' better attitude towards mathematics have been shown in previous studies.

From figure 1 above it is visible that there are some quite strong correlations with some of the variables. For instance the variables of deep learning `OrganizedStudying` and `TimeManagement` have a correlation coefficient of $\rho \approx 0,7$ as the attitude of a student correlates with their points with a coefficient $\rho \approx 0,43$. Correlations are analyzed more thoroughly below figure 2.

```{r Corrplot, out.height="80%", out.width="80%"}
corrplot(cor(learning2014[, -1]), type = "lower", diag = T, tl.col='black', tl.srt = 0.1, tl.offset = 1, 
				 title = "Fig. 2: Correlations of learning2014 data")
```

From figure 2 it can be concluded that our variables correlate the most with the variables they comprise the original variables (Deep_adj) with. Also notable is the correlation with the students' attitude and points from the final exam being approximately `r round(cor(learning2014$Points, learning2014$Attitude), digits=2)`. This means better attitude associates with better performance in the exam. 

### Logregiä varten plotteja joissa ei ole points toisena Gender differnces

Katsotaan lyhyesti läpi sukupuolten väliset erot log reg varten. fig 1 perusteella mielenkiintoisimmat ovat `OrganizedStudying` ja `Attitude`. 

```{r plotteja, fig.asp=.4}
multiplot <- function(..., plotlist=NULL, file, cols=1, layout=NULL) {
  library(grid)

  # Make a list from the ... arguments and plotlist
  plots <- c(list(...), plotlist)

  numPlots = length(plots)

  # If layout is NULL, then use 'cols' to determine layout
  if (is.null(layout)) {
    # Make the panel
    # ncol: Number of columns of plots
    # nrow: Number of rows needed, calculated from # of cols
    layout <- matrix(seq(1, cols * ceiling(numPlots/cols)),
                    ncol = cols, nrow = ceiling(numPlots/cols))
  }

 if (numPlots==1) {
    print(plots[[1]])

  } else {
    # Set up the page
    grid.newpage()
    pushViewport(viewport(layout = grid.layout(nrow(layout), ncol(layout))))

    # Make each plot, in the correct location
    for (i in 1:numPlots) {
      # Get the i,j matrix positions of the regions that contain this subplot
      matchidx <- as.data.frame(which(layout == i, arr.ind = TRUE))

      print(plots[[i]], vp = viewport(layout.pos.row = matchidx$row,
                                      layout.pos.col = matchidx$col))
    }
  }
}
p1 <- ggplot(learning2014, aes(x=Gender, y=Attitude, fill=Gender))+
	geom_boxplot()+
	geom_jitter(position=position_jitter(0.1), alpha=.4)

p2 <- ggplot(learning2014, aes(x=Gender, y=OrganizedStudying, fill=Gender))+
	geom_boxplot()+
	geom_jitter(position=position_jitter(0.1), alpha=.4)

multiplot(p1, p2, cols=2)
```



## Methods



## Results

### Summaries and visualizations

### Interpretations

### Validation analysis



## Conclusions and discussion
